\Ace: An Actively-Typed Compilation Environment\\for High-Performance Computing

### 2012-09-05 ipdps13/clq-ipdps13.tex
IPDPS13: Cleaned up intro beginning

\section{Introduction}
Researchers design and implement novel parallel programming abstractions\- on a regular basis. Each abstraction aims to make some class of developers more productive by making appropriate trade-offs between performance, portability, verifiability and ease-of-use. It can be observed that the most widely-adopted abstractions to date are those that been implemented as libraries that can be easily deployed within existing, widely-used languages. Abstractions that require introducing new primitive constructs, changing the semantics of existing constructs in particular contexts, or controlling compilation in a fine-grained manner cannot take this approach. This has led to a proliferation of specialized programming languages, each designed around a few privileged abstractions. Unfortunately, such languages have not been widely adopted by practitioners.

Computer-aided simulations and data analysis techniques have transformed science and engineering. Surveys show that scientists and engineers now spend up to 40\% of their time writing software, generally individually or in small groups \cite{howison2011scientific}\cite{hannay2009scientists}. Most of this software targets conventional desktop hardware, but about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}. In recent years, GPUs and other kinds of coprocessors have emerged as powerful platforms for high-performance computing as well.

Many scientists and engineers prefer high-level languages like MATLAB, Python, R and Perl \cite{nguyen2010survey}, due to their powerful domain-specific libraries and their focus on productivity and flexibility. These languages are typically interpreted rather than compiled and generally do not feature static type systems, relying instead on run-time (``dynamic'') type lookup and indirection. Although this can increase the flexibility of the language, these features also negatively impact  performance. Along code paths where performance appears to be a bottleneck, developers typically turn to traditional statically-typed low-level languages like C and Fortran \cite{4222616}. These languages require explicit type annotations on variables and give users explicit control over data layout and heap allocation. Although this makes writing programs more tedious and error-prone, it can also significantly  improve performance, particularly if attributes of the target architecture (e.g. caching behavior) are considered.

While a sequential reimplementation of a critical code path using a low-level language will produce a modest speedup, more significant speedups on modern hardware require parallelizing over many processor cores and, on massively-parallel machines, many interconnected nodes. Despite well-known difficulties, low-level approaches that use a form of shared-memory  multithreading (e.g. pthreads, OpenMP) paired with explicit message passing between nodes (typically using MPI) remain widely used \cite{4222616}\cite{basili2008understanding}.

>>>>

\section{Introduction}

Computer-aided simulation and data analysis techniques have transformed science and engineering. Surveys show that scientists and engineers now spend up to 40\% of their time writing software, generally individually or in small groups \cite{howison2011scientific}\cite{hannay2009scientists}. Most of this software targets desktop hardware, but about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}. To fully harness the power of modern computing resources, however, these so-called {\em professional end-user developers} \cite{segal2007some} must write parallel programs.

Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, and statically-typed, low-level languages like C and Fortran paired with low-level abstractions like pthreads and MPI for performance-critical tasks. These low-level languages and abstractions are notoriously tedious and difficult to use correctly, even for expert programmers. CITATION?

The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important factors like performance, portability, verifiability and ease-of-use. Unfortunately, professional end-users have not adopted any of these {\em en masse}, and users have become  skeptical of new approaches. This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study by Basili et al. \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by those in the research community working to advance the practice of scientific and high-performance computing.

We begin in Section 2 by developing a set of design and adoption criteria for new languages and abstractions based on prior empirical studies of these {\it professional end-user developers} as well as our observations of characteristics common to successful projects in the past.

Scientists and engineers often use high-level scripting languages like MATLAB, Python, R and Perl \cite{nguyen2010survey}, due to their powerful domain-specific libraries and their focus on ease-of-use and flexibility. Scripting languages are typically interpreted rather than compiled and generally do not feature static type systems, relying instead on run-time  type lookup and indirection. Although this can increase the flexibility of the language, these features negatively impact  performance and code verifiability.

Along code paths where performance appears to be a bottleneck, professional end users typically turn to statically-typed low-level languages like C and Fortran \cite{4222616}. These languages are statically and explicitly typed and require users to explicitly control data layout and heap allocation. Although this makes writing programs more tedious and error-prone, it can significantly improve performance, particularly if attributes of the target architecture (e.g. caching behavior) and the problem domain are carefully considered.

Although even a sequential reimplementation of a critical code path using a low-level language will produce a modest speedup, more significant speedups on modern hardware require parallelizing over many processor cores and, on massively-parallel machines, many interconnected nodes. Despite well-known difficulties, low-level approaches that use a form of shared-memory concurrency (e.g. pthreads, OpenMP) paired with explicit message passing between nodes (typically using MPI) are the most widely-used \cite{4222616}\cite{basili2008understanding}.

Researchers design and implement novel parallel programming abstractions\- on a regular basis. Each abstraction aims to make some class of developers more productive by making appropriate trade-offs between performance, portability, verifiability and ease-of-use. It can be observed that the most widely-adopted abstractions to date are those that been implemented as libraries that can be easily deployed within existing, widely-used languages. Abstractions that require introducing new primitive constructs, changing the semantics of existing constructs in particular contexts, or controlling compilation in a fine-grained manner cannot take this approach. This has led to a proliferation of specialized programming languages, each designed around a few privileged abstractions. Unfortunately, such languages have not been widely adopted by practitioners.

slash
